Efficient High-Level Abstractions for Web Programming

### 2013-07-15 papers/gpce2013/gpce2013.tex
update papers

\usepackage{amsmath}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage[unicode=true,hidelinks]{hyperref}
\usepackage{graphicx}

@@

\titlebanner{banner above paper title}        % These are ignored unless
\preprintfooter{short description of paper}   % 'preprint' option specified.

@@

\begin{abstract}
Writing large Web applications is known to be difficult. One challenge comes from the fact that the application's
logic is scattered into heterogeneous clients and servers, making it difficult to share code between both sides or to
move code from one side to the other. Another challenge is performance: while Web applications rely on ever more
code on the client-side, they may run on smart phones with little hardware capabilities. These two challenges raise
the following problem: how to benefit from high-level languages and libraries making code complexity easier to manage
and abstracting over the clients and servers differences without trading this engineering comfort for performance?
This article presents high-level abstractions defined as deep embedded DSLs in Scala, that can (1) generate efficient
code leveraging the target platform characteristics, (2) be shared between clients and servers. Though code written
with our DSL has a high level of abstraction our benchmark on a real world application reports that it runs as fast
as hand tuned low-level JavaScript code.
\end{abstract}

@@

Web applications are attractive because they require no installation or deployment steps on clients and enable large
scale collaborative experiences. However, writing large Web applications is known to be
difficult~\cite{Mikkonen08_SpaghettiJs,Preciado05_RIAMethodologyNecessity}. One challenge comes from the fact
that the business logic is scattered into heterogeneous client-side and server-side
environments~\cite{Echeverria09_RIA,Kuuskeri09_PartitioningClientServer}. This gives less flexibility in the
engineering process and requires a higher maintenance effort: there is no way to move a piece of code targeting
the server-side to target the client-side, the code has to be rewritten. Even worse, logic parts that run on both
client-side and server-side need to be duplicated. For instance, HTML fragments may be built from the server-side
when a page is requested by a client, but they may also be built from the client-side to perform an incremental
update subsequent to an user action. How could developers write HTML fragment definitions once and render them on
both client-side and server-side?

The more interactive the application is, the more logic needs to be duplicated between the server-side and the
client-side, and the higher is the complexity of the client-side code. Developers use libraries and frameworks to get
high-level abstractions on client-side, making their code easier to reason about and to maintain, but also making
their code run less efficiently (due to \emph{abstraction penalty}).

Performance is a primary concern in Web applications, because they are expected to run on a broad range of devices,
from the powerful desktop personal computer to the less powerful smart phone. “Every 100~ms delay costs 1\% of
sales”, said Amazon in 2006.

Using the same programming language on both server-side and client-side could improve the software engineering
process by enabling code reuse between both sides. Incidentally, the JavaScript language -- which is currently the
most supported the action language on Web clients -- can be used on server-side, and an increasing number of
programming languages or compiler back-ends can generate JavaScript code (\eg Java/GWT~\cite{Chaganti07_GWT},
SharpKit\footnote{\href{http://sharpkit.net}{http://sharpkit.net}}, Dart~\cite{Griffith11_Dart},
Kotlin\footnote{\href{http://kotlin.jetbrains.org/}{http://kotlin.jetbrains.org/}},
ClojureScript~\cite{McGranaghan11_ClojureScript}, Fay\footnote{\href{http://fay-lang.org/}{http://fay-lang.org/}},
Haxe~\cite{Cannasse08_HaXe} or Opa\footnote{\href{http://opalang.org/}{http://opalang.org/}}).

@@

Generating efficient code for heterogeneous platforms is hard to achieve in an extensible way: the translation of
common abstractions like collections into their native counterpart (JavaScript arrays on client-side and standard
library's collections on server-side) may be hard-coded in the compiler, but that approach would not scale to handle
all the abstractions a complete application may use (\eg HTML fragment definitions, form validation rules, or even
some business data type that may be represented differently).

On one hand, for engineering reasons, developers want to write Web applications using a single high-level language,
abstracting over the target platforms differences and reducing code complexity. But on the other hand, for
performance reasons, they want to keep control on the way their code is compiled to each target platform. How to
solve this dilemma?

Compiled domain-specific embedded languages~\cite{Elliott2003_Compiling} allow the definition of domain-specific
languages (DSLs) as libraries on top of a host language, and to compile them to a target platform. The deep embedding
gives the opportunity to control the code generation scheme for a given abstraction and target platform.

@@

\begin{itemize}
 \item we define high-level abstractions, typically used in Web programming, that generate low-level code leveraging
native Web browsers APIs;
 \item in the case of abstractions shared between servers and clients, we specialize the code generation in order to
leverage the target platform native environments.
\end{itemize}

@@

The remainder of this paper is organized as follows. The next section introduces existing approaches for defining
cross-compiling languages and presents the framework we used to define our DSLs. Sections \ref{contribution} and
\ref{contribution-shared} present our contribution. Section \ref{validation} compares our solution to common
approaches. Section \ref{discussion} concludes.

\section{Background}

\subsection{Language Engineering Processes}

This section presents different approaches for defining cross-platform programming languages.

\begin{figure}
\begin{center}
\includegraphics[width=6cm]{langs.pdf}
\end{center}
\caption{Language engineering processes}
\label{langs}
\end{figure}

@@

The first approach for defining a cross-platform language consists in hard-coding, in the compiler, the
code generation scheme of each language feature to each target platform. Figure \ref{langs} (a) depicts this
process. In order to support a feature related to a specific domain, the whole compiler pipeline (parser, code
generator, \etc) may have to be adapted. This approach gives \emph{fat} languages because a lot of concepts are
defined at the language level: general programming concepts such as naming, functions, classes, as well as more
domain-specific concepts such as HTML fragment definition. Examples of such languages for Web programming are
Links~\cite{Cooper07_Links}, Opa and Dart~\cite{Griffith11_Dart}. These languages are difficult to extend because
each concept is defined in the compiler, and modifying a compiler requires a high effort. Furthermore, these
languages also require to support common programming abstraction and composition mechanisms, as general purpose
languages do. So they usually try to re-invent the features of general purpose languages, that’s why we argue that
this approach for defining programming languages is difficult to scale: for every specific problem you would have to
rewrite a full-featured programming language besides addressing the concepts specific to the problem domain.

@@

Another approach consists in defining several independent domain-specific languages~\cite{Van00_DSL}, each one
focusing on concerns specific to a given problem domain, and then to combine all the source artifacts written with
these language into one executable program, as shown in figure~\ref{langs} (b). Defining such languages requires a
minimal effort compared to the previous approach because each language has a limited set of features. On the other
hand, it is difficult to have interoperability between DSLs (missing ref). \cite{Groenewegen08_WebDSL} gave an
example of such a domain-specific language for defining Web applications.

@@

Alternatively, one can define concepts relative to a specific domain as a library on top of a thin general purpose
language (it is also referred to as a domain-specific \emph{embedded} language~\cite{Hudak96_DSEL}). Figure
\ref{langs} (c) depicts this approach. The general purpose language is used as a host language and does not need to
be modified if a new concept is introduced, because concepts are defined as pure libraries. On the other hand, the
syntax of the DSLs is limited by the syntax flexibility of the host language. Furthermore, this approach gives no
opportunity to translate a concept efficiently according to the target platform characteristics because concepts are
defined as libraries and are translated by the compiler, which has no domain-specific knowledge. Examples of
languages following this approach are Java/GWT, Kotlin, HaXe and SharpKit.

@@

The last approach, shown in figure \ref{langs} (d), can be seen as a middle-ground between the two previous
approaches: DSLs are embedded in a host language but use a code generation process. This approach share the same
benefits and limits as embedded DSLs for defining language units. The code generation process is specific to each DSL
and gives the opportunity to perform domain-specific optimizations. In other words deeply embedded DSLs bring
domain-specific knowledge to the compiler.

(Summary explaining why we choose the deep embedded DSLs approach)

\subsection{Lightweight Modular Staging}

@@

\section{High-Level Abstraction for Client-Side Code: Selectors}
\label{contribution}

@@

\begin{figure}
\begin{center}
\begin{tabular}{| l | p{3cm} |}
\hline
Function & Description \\
\hline
\code{querySelector(s)} & First element matching the CSS selector \code{s} \\
\hline
\code{getElementById(i)} & Element which attribute \code{id} equals to \code{i} \\
\hline
\code{querySelectorAll(s)} & All elements matching the CSS selector \code{s} \\
\hline
\code{getElementsByTagName(n)} & All elements of type \code{n} \\
\hline
\code{getElementsByClassName(c)} & All elements which \code{class} attribute contains \code{c} \\
\hline
\end{tabular}
\end{center}
\caption{Standard selectors API}
\label{selectors-api}
\end{figure}

The \code{querySelector} and \code{querySelectorAll} are the most general functions while the others handle special
cases. For the developer it is not convenient to have to master several functions performing similar tasks. In fact,
most JavaScript developers use the jQuery
library~\cite{Bibeault08_jQuery}\footnote{According to \href{http://trends.builtwith.com/javascript}{
http://trends.builtwith.com/javascript}, jQuery is used by more than
40\% of the top million sites} that provides only one high-level function to search for elements.
Listings \ref{vanilla-selectors} and \ref{jquery-selectors} show two equivalent JavaScript programs performing
element searches, the first one using the native APIs and the second one using jQuery.

\begin{figure}
\begin{lstlisting}[language=JavaScript,label=vanilla-selectors,caption=Searching elements in plain JavaScript]
function getWords() {
  var form = document.getElementById('add-user');
  var sections =
    form.getElementsByTagName('fieldset');
  var results = [];
  for (var i = 0 ; i < sections.length ; i++) {
    var words = sections[i]
      .getElementsByClassName('word');
    results[i] = words;
  }
  return results
}
\end{lstlisting}
\end{figure}

\begin{figure}
\begin{lstlisting}[language=JavaScript,label=jquery-selectors,caption=Searching elements in jQuery]
function getWords() {
  var form = $('#add-user');
  var sections = $('fieldset', form);
  return sections.map(function () {
    return $('.word', this)
  })
}
\end{lstlisting}
\end{figure}

jQuery provides an API that is simpler to master because it has less functions, but by doing so it can not benefit
from the performance of the browser’s implementation of specialized search functions (\code{getElementById},
\code{getElementsByTagName} and \code{getElementsByClassName}).

\begin{figure}
\begin{lstlisting}[label=js-scala-selectors,caption=Searching elements in js-scala]
def getWords() = {
  val form = document.find("#add-user")
  val sections = form.findAll("fieldset")
  sections map (_.findAll(".word"))
}
\end{lstlisting}
\end{figure}

Listing~\ref{js-scala-selectors} shows how to implement listing~\ref{jquery-selectors} using js-scala. We
provide two functions for searching elements: \code{find} to find the first element matching a selector and
\code{findAll} to find all the matching elements. During the first evaluation step, these functions try to analyze
the selector that is passed as parameter and, when appropriate, produce code using the specialized  API, otherwise
they produce  code using \code{querySelector} and \code{querySelectorAll}. As a result,
listing~\ref{js-scala-selectors} generates a JavaScript program identical to listing~\ref{vanilla-selectors}: the
high-level abstraction (the \code{find} and \code{findAll} functions) exist only in the initial program, not in the
final JavaScript program.

\begin{figure}
\begin{lstlisting}[label=selector-impl,caption=Selectors optimization]
def find(receiver: Rep[Selector],
         selector: Rep[String]) =
  getConstIdCss(selector) match {
    case Some(id) if receiver == document =>
      DocumentGetElementById(Const(id))
    case _ =>
      SelectorFind(receiver, selector)
  }
\end{lstlisting}
\end{figure}

Listing \ref{selector-impl} shows the implementation of the \code{find} function producing different IR nodes
according to the selector passed as parameter. The \code{getConstIdCss} function analyzes the selector: if it
is a constant \code{String} value containing a CSS ID selector, it returns the value of the identifier. If the
\code{find} function is applied to the \code{document} and to an ID selector, it returns a
\code{DocumentGetElementById} IR node (that is translated to a \code{document.getElementById} call by the code
generator), otherwise it returns a \code{SelectorFind} IR node (that is translated to a \code{querySelector}
call). (FIXME Add a ref to the spec of CSS selectors?)

@@

\section{High-Level Abstractions Shared on Clients and Servers}
\label{contribution-shared}

\subsection{Monads Sequencing}

Null references are a known source of problems in programming languages~\cite{Hoare09_Null,Nanda09_Null}. For
example, consider listing \ref{null-unsafe} finding a particular widget in the page and then a particular
button within the widget. The native \code{querySelector} method returns \code{null} if no node matched the given
selector in the document. If we run this code in a page where the widget is not present, it will throw an error
and stop further JavaScript execution. Defensive code can be written to handle \code{null} references, but leads to
very cumbersome code, as shown in listing \ref{null-defensive}.

\begin{figure}
\begin{lstlisting}[language=JavaScript,label=null-unsafe,caption=Unsafe code]
var loginWidget =
  document.querySelector("div.login");
var loginButton =
  loginWidget.querySelector("button.submit");
loginButton.addEventListener("click", handler);
\end{lstlisting}
\end{figure}

\begin{figure}
\begin{lstlisting}[language=JavaScript,label=null-defensive,caption=Defensive programming to handle null references]
var loginWidget =
  document.querySelector("div.login");
if (loginWidget !== null) {
  var loginButton =
    loginWidget.querySelector("button.submit");
  if (loginButton !== null) {
    loginButton.
      addEventListener("click", handler);
  }
}
\end{lstlisting}
\end{figure}

Some programming languages encode optional values with a monad (\eg \code{Maybe} in Haskell and \code{Option} in
Scala). In that case, sequencing over the monad encodes optional value dereferencing. If the language supports a
convenient syntax for monad sequencing, it brings a convenient syntax for optional value dereferencing, alleviating
developers from the burden of defensive programming.

Listing \ref{null-js-scala} implements in js-scala a program equivalent to listing \ref{null-defensive}. The
\code{for} notation is used to sequence computations over optional values that are encoded with a monad. The
\code{find} function returns a \code{Rep[Option[Element]]} value, that can either be a \code{Rep[Some[Element]]} (if
an element was found) or a \code{Rep[None.type]} (if no element was found). The \code{for} expression contains a
sequence of statements that are executed in order, as long as the previous statement returned a
\code{Rep[Some[Element]]} value.

\begin{figure}
\begin{lstlisting}[label=null-js-scala,caption=Handling null references in js-scala]
for {
  loginWidget <- document.find("div.login")
  loginButton <- loginWidget.find("submit.button")
} loginButton.on(Click)(handler)
\end{lstlisting}
\end{figure}

Such a monadic API brings both safety and expressiveness to developers manipulating optional values but usually
involves the creation of an extra container object holding the optional value. In our case, the monadic API is
used in the initial program but generates code that does not wrap values in container objects but instead checks if
they are \code{null} or not when dereferenced. So the extra container object exists only in the initial program and
is removed during code generation: listing \ref{null-js-scala} produces a code equivalent to listing
\ref{null-defensive}.

\begin{figure}
\begin{lstlisting}[caption=JavaScript code generator for null references handling DSL,label=option-codegen]
override def emitNode(sym: Sym[Any], rhs: Def[Any]) =
  rhs match {
    case OptionIsEmpty(o) =>
      emitValDef(sym, q" $o === null")
    case OptionForeach(o, b) =>
      stream.println(q"if ($o !== null) {")
      emitBlock(b)
      stream.println("}")
    case _ =>
      super.emitNode(sym, rhs)
  }
\end{lstlisting}
\end{figure}

@@

Last but not least, we make this abstraction available on server-side by writing a code generator similar to the
JavaScript code generator, but targeting Scala (FIXME Show the code generator?). So the same abstraction is
efficiently translated on both server and client sides.

\subsection{DOM Fragments Definition}

@@

\begin{figure}
\begin{lstlisting}[language=JavaScript,caption=JavaScript DOM API,label=dom-api]
var articleUi = function (article) {
  var div = document.createElement('div');
  div.setAttribute('class', 'article');
  var span = document.createElement('span');
  var name =
    document.createTextNode(article.name + ': ');
  span.appendChild(name);
  div.appendChild(span);
  var strong = document.createElement('strong');
  var price = document.createTextNode(article.price);
  strong.appendChild(price);
  div.appendChild(strong);
  return div
};
\end{lstlisting}
\end{figure}

\begin{figure}
\begin{lstlisting}[caption=Scala XML API,label=scala-xml-api]
def articleUi(article: Article) =
  <div class="article">
    <span>{ article.name + ": " }</span>
    <strong>{ article.price }</strong>
  </div>
\end{lstlisting}
\end{figure}

@@

Instead, we provide a common high-level DSL for defining HTML fragments and we make this DSL generate code leveraging
native environments. Listing \ref{forest} shows how to implement our example with our DSL. The \code{el} function
defines an HTML element, eventually containing attributes and children elements. Any children of an element that is
not an element itself is converted into a text node. The children elements of an element can also be obtained
dynamically from a collection, as shown in listing \ref{forest-loops}.

>>>>

\usepackage{amsmath}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage[unicode=true,hidelinks]{hyperref}
\usepackage{graphicx}
\usepackage{moresize}

@@

%\titlebanner{banner above paper title}        % These are ignored unless
%\preprintfooter{short description of paper}   % 'preprint' option specified.

@@

\begin{abstract}
%\vspace{0.2cm}

Writing large Web applications is known to be difficult. One challenge comes from the fact that the
application's logic is scattered into heterogeneous clients and servers, making it difficult to
share code between both sides or to move code from one side to the other. Another challenge is
performance: while Web applications rely on ever more code on the client-side, they may run on smart
phones with limited hardware capabilities. These two challenges raise the following problem: how to
benefit from high-level languages and libraries making code complexity easier to manage and
abstracting over the clients and servers differences without trading this engineering comfort for
performance? This article presents high-level abstractions defined as deep embedded DSLs in Scala,
that can (1) generate efficient code leveraging the target platform characteristics, (2) be shared
between clients and servers. Though code written with our DSL has a high level of abstraction our
benchmark on a real world application reports that it runs as fast as hand tuned low-level
JavaScript code.
\end{abstract}

@@

%\vspace{0.2cm}

Web applications are attractive because they require no installation or deployment steps on clients
and enable large scale collaborative experiences. However, writing large Web applications is known
to be difficult~\cite{Mikkonen08_SpaghettiJs,Preciado05_RIAMethodologyNecessity}. One challenge
comes from the fact that the business logic is scattered into heterogeneous client-side and
server-side environments~\cite{Echeverria09_RIA,Kuuskeri09_PartitioningClientServer}. This gives
less flexibility in the engineering process and requires a higher maintenance effort: there is no
way to move a piece of code targeting the server-side to target the client-side -- the code has to
be rewritten. Even worse, logic parts that run on both client-side and server-side need to be
duplicated. For instance, HTML fragments may be built from the server-side when a page is requested
by a client, but they may also be built from the client-side to perform an incremental update
subsequent to a user action. How could developers write HTML fragment definitions once and render
them on both client-side and server-side?

The more interactive the application is, the more logic needs to be duplicated between the
server-side and the client-side, and the higher is the complexity of the client-side code.
Developers can use libraries and frameworks to get high-level abstractions on client-side, making
their code easier to reason about and to maintain, but also making their code run less efficiently
due to \emph{abstraction penalty}.

Performance is a primary concern in many Web applications, because they are expected to run on a
broad range of devices, from the powerful desktop personal computer to the less powerful smart
phone~\cite{Souders08_Perf,Huang10_Perf}.

Using the same programming language on both server-side and client-side could improve the software
engineering process by enabling code reuse between both sides. Incidentally, the JavaScript language
-- which is currently the most supported action language on Web clients -- can be used on
server-side. Conversely, an increasing number of programming languages or compiler back-ends can
generate JavaScript code (\eg{} Java/GWT~\cite{Chaganti07_GWT},
SharpKit\footnote{\href{http://sharpkit.net}{http://sharpkit.net}}, Dart~\cite{Griffith11_Dart},
Kotlin\footnote{\href{http://kotlin.jetbrains.org/}{http://kotlin.jetbrains.org/}},
ClojureScript~\cite{McGranaghan11_ClojureScript},
Fay\footnote{\href{http://fay-lang.org/}{http://fay-lang.org/}},
Haxe~\cite{Cannasse08_HaXe} or Opa\footnote{\href{http://opalang.org/}{http://opalang.org/}}).

@@

Generating efficient code for heterogeneous platforms is hard to achieve in an extensible way: the
translation of common abstractions like collections into their native counterpart (JavaScript arrays
on client-side and standard library's collections on server-side) may be hard-coded in the compiler,
but that approach would not scale to handle all the abstractions a complete application may use
(\eg{} HTML fragment definitions, form validation rules, or even some business data type that may be
represented differently).

On one hand, for engineering reasons, developers want to write Web applications using a single
high-level language, abstracting over the target platforms differences and reducing code complexity.
But on the other hand, for performance reasons, they want to keep control on the way their code is
compiled to each target platform. We propose to solve this dilemma by providing high-level
abstractions in compiled domain-specific embedded languages (DSELs)~\cite{Hudak96_DSEL,
Elliott2003_Compiling}. Compiled DSELs allow the definition of domain-specific languages (DSLs) as
libraries on top of a host language, and to compile them to a target platform. Their deep embedding
gives the opportunity to control the code generation scheme for a given abstraction and target
platform.

@@

\begin{itemize}
 \item we define high-level abstractions, typically used in Web programming, that generate low-level
code leveraging native Web browser APIs;
 \item in the case of abstractions shared between servers and clients, we specialize the code
generation in order to leverage the target platform native environments.
\end{itemize}

@@

The remainder of this paper is organized as follows. The next section overviews the
existing approaches defining high-level languages for Web programming. Section \ref{sec:lms}
presents the framework we used to define our DSLs. Section \ref{sec:contribution} presents our
contribution. Section \ref{sec:validation} compares our solution to common approaches. Section
\ref{sec:discussion} discusses our results and section \ref{sec:conclusion} concludes.

\section{Related Work}
\label{sec:related-work}

%\vspace{0.1cm}

We classified existing approaches providing high-level abstractions for Web programming in four
categories, as shown in figure \ref{fig:langs}.

\begin{figure}[htb]
\begin{center}
\includegraphics[width=7.5cm]{langs.pdf}
\end{center}
\caption{Language engineering processes}
\vspace{-0.5cm}
\label{fig:langs}
\end{figure}

@@

\enlargethispage{0.3cm}
%\vspace{0.1cm}

The first approach for defining a cross-platform language consists in hard-coding, in the compiler,
the code generation scheme of each language feature to each target platform. Figure \ref{fig:langs}
(a) depicts this process. In order to support a feature related to a specific domain, the whole
compiler pipeline (parser, code generator, \etc) may have to be adapted. This approach gives
\emph{fat} languages because a lot of concepts are defined at the language level: general
programming concepts such as naming, functions, classes, as well as more domain-specific concepts
such as HTML fragment definition. Thus, implementing a fat language may require a high effort.
Examples of such languages for Web programming are Links~\cite{Cooper07_Links}, Opa and
Dart~\cite{Griffith11_Dart}.

@@

%\vspace{0.1cm}

Another approach consists in defining several independent domain-specific
languages~\cite{Van00_DSL}, each one focusing on concerns specific to a given problem domain, and
then to combine all the source artifacts written with these language into one executable program, as
shown in figure~\ref{fig:langs} (b). Defining such languages requires a minimal effort compared to
the previous approach because each language has a limited set of features. On the other hand, it is
difficult to have interoperability between DSLs. \cite{Groenewegen08_WebDSL} gave an example of such
a domain-specific language for defining Web applications.

@@

\enlargethispage{0.3cm}

%\vspace{0.1cm}

Alternatively, one can define concepts relative to a specific domain as a library on top of a thin
general purpose language (it is also referred to as a domain-specific \emph{embedded}
language~\cite{Hudak96_DSEL}). Figure \ref{fig:langs} (c) depicts this approach. Defining such a
library requires minimal effort (though the syntax of the DSL is limited by the syntax flexibility
of the host language) and several DSLs can interoperate freely within the host language. However,
this approach gives no opportunity to efficiently translate a concept according to the target
platform characteristics because the compiler has no domain-specific knowledge (though some
compilers hard-code the translation of some common abstractions such as arrays to leverage the
target platform characteristics). Examples of languages following this approach are Java/GWT,
Kotlin, HaXe and SharpKit.

@@

%\vspace{0.1cm}
The last approach, shown in figure \ref{fig:langs} (d), can be seen as a middle-ground between the
two previous approaches: DSLs are embedded in a host language but use a code generation process.
This approach shares the same benefits and limitations as embedded DSLs for defining language units.
However, the code generation process is specific to each DSL and gives the opportunity to perform
domain-specific optimizations. In other words deeply embedded DSLs bring domain-specific knowledge
to the compiler. Js-scala~\cite{Kossakowski12_JsDESL} is an example of deeply embedded DSL in Scala
for Web programming. It makes it possible to produce JavaScript programs from Scala code that uses
basic language concepts like arrays and control structures (\code{if} and \code{while}) as well as
mechanisms specific to the Scala compiler, like delimited continuations to handle asynchronous
computations. Though the initial authors of js-scala showed how to use the native browser APIs in an
untyped fashion, they did not show how to build high-level abstractions for Web programming
generating efficient code leveraging the browser APIs and did not show how to share such
abstractions between server and client sides (they achieved code sharing, yet on the server-side
their DSL was not compiled but interpreted). Our paper addresses these limitations.

\section{Lightweight Modular Staging}
\label{sec:lms}

This section gives background material on the framework used to define js-scala.

@@

\section{Efficient High-Level Abstractions for Web Programming}
\label{sec:contribution}

This section presents some tasks typically performed in Web applications, either on client-side or
server-side or on both, generalizes them in terms of high-level abstractions, and shows how they are
implemented in js-scala to generate efficient code.

\subsection{Selectors API}
\label{sec:selectors}

@@

\begin{figure}[htb]
\begin{center}
\begin{tabular}{| l | p{3cm} |}
\hline
Function & Description \\
\hline
\code{querySelector(s)} & First element matching the CSS selector \code{s} \\
\hline
\code{getElementById(i)} & Element which attribute \code{id} equals to \code{i} \\
\hline
\code{querySelectorAll(s)} & All elements matching the CSS selector \code{s} \\
\hline
\code{getElementsByTagName(n)} & All elements of type \code{n} \\
\hline
\code{getElementsByClassName(c)} & All elements which \code{class} attribute contains \code{c} \\
\hline
\end{tabular}
\end{center}
\caption{Standard selectors API. The \code{querySelector} and \code{querySelectorAll} are the most
general functions while the others handle special cases.}
\label{selectors-api}
\end{figure}

\begin{figure}[htb]
\begin{lstlisting}[language=JavaScript,label=vanilla-selectors,caption=Searching elements using the
native selectors API]
function getWords() {
  var form = document.getElementById('add-user');
  var sections =
    form.getElementsByTagName('fieldset');
  var results = [];
  for (var i = 0 ; i < sections.length ; i++) {
    var words = sections[i]
      .getElementsByClassName('word');
    results[i] = words;
  }
  return results
}
\end{lstlisting}
\end{figure}

Listing \ref{vanilla-selectors} gives an example of use of various functions from the native
selectors API to retrieve a list of input fields within a form. The existence of several
specialized functions in the API makes it possible to write efficient code, but forces users to
think at a low abstraction level.

\begin{figure}[htb]
\begin{lstlisting}[language=JavaScript,label=jquery-selectors,caption=Searching elements in jQuery]
function getWords() {
  var form = $('#add-user');
  var sections = $('fieldset', form);
  return sections.map(function () {
    return $('.word', this)
  })
}
\end{lstlisting}
\end{figure}

A high-level abstraction for searching elements in a document could be just one function
finding all elements matching a given CSS selector. In fact, most JavaScript
developers\footnote{According to
\href{http://trends.builtwith.com/javascript}{http://trends.builtwith.com/javascript}, jQuery is
used by more than 40\% of the top million sites.} use the jQuery library~\cite{Bibeault08_jQuery}
that actually provides only one function to search for elements. Listing \ref{jquery-selectors}
shows an equivalent JavaScript program as listing \ref{vanilla-selectors}, but using jQuery. The
code is both shorter and simpler, thanks to its higher-level of abstraction. jQuery provides an API
that is simpler to master because it has less functions, but this benefit comes at the price of a
decrease in runtime performances.

Instead, we propose a solution that has a high-level API but generates JavaScript code using the
specialized native API, when possible, in order to get both engineering comfort and performances.
We achieve this by analyzing, during the first evaluation step, the selector that is
passed as parameter and, when appropriate, by producing JavaScript code using the specialized API,
and otherwise producing code using \code{querySelector} and \code{querySelectorAll}.

\begin{figure}[htb]
\begin{lstlisting}[label=selector-impl,caption=Selectors optimization]
def find(selector: Rep[String]) =
  getConstIdCss(selector) match {
    case Some(id) if receiver == document =>
      DocumentGetElementById(Const(id))
    case _ =>
      SelectorFind(receiver, selector)
  }
\end{lstlisting}
\end{figure}

Our API has two functions: \code{find} to find the first element matching a selector and
\code{findAll} to find all the matching elements. Listing \ref{selector-impl} gives the
implementation of the \code{find} function. It is a Scala function that returns an IR node
representing the JavaScript computation that will search the element in the final program. The
\code{getConstIdCss} function analyzes the selector: if it is a constant \code{String} value
containing a CSS ID selector, it returns the value of the identifier. So, if the \code{find}
function is applied to the \code{document} and to an ID selector, it returns a
\code{DocumentGetElementById} IR node (that is translated to a \code{document.getElementById} call
by the code generator), otherwise it returns a \code{SelectorFind} IR node (that is translated to a
\code{querySelector} call).

@@

\begin{figure}[htb]
\begin{lstlisting}[label=js-scala-selectors,caption=Searching elements in js-scala]
def getWords() = {
  val form = document.find("#add-user")
  val sections = form.findAll("fieldset")
  sections map (_.findAll(".word"))
}
\end{lstlisting}
\end{figure}

\begin{figure}[htb]
\begin{center}
\includegraphics[width=5cm]{ir.pdf}
\end{center}
\caption{Intermediate representations returned by the evaluation of (a)
\code{document.find("\#add-user button")} and (b) \code{document.find("\#add-user")}}
\label{fig:ir}
\end{figure}

\vspace{1 cm}

Figure \ref{fig:ir} shows the IRs returned by the evaluation of \small{
$document\\.find("\#add-user button")$} \normalsize{and}  \small{$ document.find("\#add-user")$}.

\normalsize{}

\sloppy
In the former case, the selector is parsed and does not match an ID selector (it is a composite
selector matching button elements within the element having the \code{add-user} id), so a
\code{SelectorFind} node is returned, then translated into a call to the general
\code{querySelector} function. In the latter case, the selector matches an ID selector so a
\code{DocumentGetElementById} node is returned, then translated into a call to the specialized
\code{getElementById} function.

Finally, listing~\ref{js-scala-selectors} shows how to implement listing~\ref{jquery-selectors} in
Scala using js-scala. The code has the same abstraction level as with jQuery, however it
generates a JavaScript program identical to listing~\ref{vanilla-selectors}: the high-level
abstractions (the \code{find} and \code{findAll} functions) exist only in the initial program, not
in the final JavaScript program.

\subsection{Monads Sequencing}
\label{sec:option}

This section presents an abstraction to handle \code{null} references and shows how this
abstraction can be shared between client and server code.

\code{null} references are a known source of problems in programming
languages~\cite{Hoare09_Null,Nanda09_Null}. For example, consider listing \ref{null-unsafe} finding
a particular widget in the page and then a particular button within the widget. The native
\code{querySelector} method returns \code{null} if no node matched the given selector in the
document. If we run this code in a page where the widget is not present, it will throw an error
and stop further JavaScript execution. Defensive code can be written to handle \code{null}
references, but leads to very cumbersome code, as shown in listing \ref{null-defensive}.

\begin{figure}[htb]
\begin{lstlisting}[language=JavaScript,label=null-unsafe,caption=Unsafe code]
var loginWidget =
  document.querySelector("div.login");
var loginButton =
  loginWidget.querySelector("button.submit");
loginButton.addEventListener("click", handler);
\end{lstlisting}
\end{figure}

\begin{figure}[htb]
\begin{lstlisting}[language=JavaScript,label=null-defensive,caption=Defensive programming to handle
null references]
var loginWidget =
  document.querySelector("div.login");
if (loginWidget !== null) {
  var loginButton =
    loginWidget.querySelector("button.submit");
  if (loginButton !== null) {
    loginButton.
      addEventListener("click", handler);
  }
}
\end{lstlisting}
\end{figure}

Some programming languages encode optional values with a monad (\eg{} \code{Maybe} in Haskell and
\code{Option} in Scala). In that case, sequencing over the monad encodes optional value
dereferencing. If the language supports a convenient syntax for monad sequencing, it brings a
convenient syntax for optional value dereferencing, alleviating developers from the burden of
defensive programming.

In our DSL, we encode an optional value of type \code{Rep[A]} using a \code{Rep[Option[A]]} value,
which can either be a \code{Rep[Some[A]]} (if there is a value) or a \code{Rep[None.type]} (if
there is no value). An optional value can be dereferenced using the \code{for} notation, as shown
in listing \ref{null-js-scala}, that implements in js-scala a program equivalent to listing
\ref{null-defensive}. The \code{find} function returns a \code{Rep[Option[Element]]}. The \code{for}
expression contains a sequence of statements that are executed in order, as long as the previous
statement returned a \code{Rep[Some[Element]]} value.

\begin{figure}[htb]
\begin{lstlisting}[label=null-js-scala,caption=Handling null references in js-scala]
for {
  loginWidget <- document.find("div.login")
  loginButton <- loginWidget.find("submit.button")
} loginButton.on(Click)(handler)
\end{lstlisting}
\end{figure}

Such a monadic API brings both safety and expressiveness to developers manipulating optional values
but usually involves the creation of an extra container object holding the optional value. In our
case, the monadic API is used in the initial program but generates code that does not wrap values in
container objects but instead checks if they are \code{null} or not when dereferenced. So the extra
container object exists only in the initial program and is removed during code generation: listing
\ref{null-js-scala} produces a code equivalent to listing~\ref{null-defensive}.

\begin{figure}[htb]
\begin{lstlisting}[caption=JavaScript code generator for null references handling
DSL,label=option-codegen,captionpos=b]
override def emitNode(sym: Sym[Any], rhs: Def[Any]) =
  rhs match {
    case OptionIsEmpty(o) =>
      emitValDef(sym, q" $o === null")
    case OptionForeach(o, b) =>
      stream.println(q"if ($o !== null) {")
      emitBlock(b)
      stream.println("}")
    case _ =>
      super.emitNode(sym, rhs)
  }
\end{lstlisting}
\end{figure}

@@

The IR nodes are not tied to the JavaScript code generator, so we are able to make this abstraction
available on server-side by writing a code generator similar to the JavaScript code generator, but
targeting Scala. So the same abstraction is efficiently translated on both server and client sides.

\subsection{DOM Fragments Definition}
\label{sec:forest}

@@

\begin{figure}[htb]
\begin{lstlisting}[language=JavaScript,caption=JavaScript DOM API,label=dom-api]
var articleUi = function (article) {
  var div = document.createElement('div');
  div.setAttribute('class', 'article');
  var span = document.createElement('span');
  var name =
    document.createTextNode(article.name + ': ');
  span.appendChild(name);
  div.appendChild(span);
  var strong = document.createElement('strong');
  var price = document.createTextNode(article.price);
  strong.appendChild(price);
  div.appendChild(strong);
  return div
};
\end{lstlisting}
\end{figure}

\begin{figure}[htb]
\begin{lstlisting}[caption=Scala XML API,label=scala-xml-api]
def articleUi(article: Article) =
  <div class="article">
    <span>{ article.name + ": " }</span>
    <strong>{ article.price }</strong>
  </div>
\end{lstlisting}
\end{figure}

@@

Our first step consists in capturing, in a high-level API, the concepts common to the JavaScript and
Scala APIs. Though they are different, both APIs define HTML elements with attributes and content.
We propose to have a function \code{el} to define an HTML element, eventually containing attributes
and children elements. Any children of an element that is not an element itself is converted into a
text node. Listing \ref{forest} shows how to implement our example with our DSL. The children
elements of an element can also be obtained dynamically from a collection, as shown in listing
\ref{forest-loops}.
